\vspace{1.0cm} \Large \bf Integrating influenza antigenic dynamics with molecular evolution

### 2013-07-24 response.tex

\textbf{We have expanded this paragraph to encompass broadly neutralizing anti-HA antibodies and antibodies to internal proteins and included the above references.}

MINOR COMMENTS:

The first paragraph says: ``A large proportion of the disease burden of influenza stems from antigenic drift; it is why vaccines remain only transiently effective.'' In fact, influenza vaccines are probably relatively ineffective for a number of reasons, and there is no correlation between vaccine match and efficacy (although the data is admittedly noisy), as discussed on page 28 of http://www.cidrap.umn.edu/cidrap/files/80/ccivi\%20report.pdf So it might be more accurate to say that ``antigenic drift is probably one reason why influenza vaccines are only transiently effective.''

\textbf{We have revised this paragraph to make it clear that antigenic drift causes efficacy to decline over time, but that there are other factors that generally limit influenza vaccine efficacy.}

@@

Could the authors elaborate in the Methods about the meaning of the lines connecting points in Figure 2 and 3? They say these represent ``mean posterior diffusions paths,'' but I'm having trouble grasping what that means.

\textbf{We apologize for the lack of clarity here.  We have included an additional paragraph in Methods under Implementation that describes the algorithm to estimate ``mean diffusion paths''.  This paragraph reads as follows:}

\textbf{``We summarize diffusion paths of viral lineages (Figure~2, Figure~3) by taking each virus and reconstructing $x$ and $y$ locations along antigenic dimensions 1 and 2 backward through time.
We use MCMC to sample tip locations, but when outputting trees sample internal node locations using a peeling algorithm as described in [22].
Thus, after the MCMC is finished we have a posterior sample of 2000 trees each tagged with estimated tip locations and internal node locations.
We post-processed each posterior tree by conducting a linear interpolation between parent-child node locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
Then, for each interval, $x$ and $y$ values are averaged across the sample of posterior trees.
We draw lines between these locations to approximate mean posterior diffusion paths.
As virus lineages coalesce backwards through time down the phylogeny these diffusion paths will also coalesce.''}

It would be helpful to have phylogenetic trees for the four clades as supporting figures, since as best I understand Figure 2 and 3 do not represent the phylogenetic relationships, but only show the influence of those relationships on the antigenic map.

Also, are the phylogenetic trees built on just HA1 or the whole HA protein?

\textbf{The trees were inferred from the whole HA gene when available, but for many viruses only HA1 was available.  The phylogenetic methods available in BEAST are able to properly deal with this sort of missing data.  We've made note of this in the text.}

### 2013-07-24 response.tex

\textbf{We apologize for the lack of clarity here.  We have included an additional paragraph in Methods under Implementation that describes the algorithm to estimate ``mean diffusion paths''.  This paragraph reads as follows:}

\textbf{``We summarize diffusion paths of viral lineages (Figure~2, Figure~3) by taking each virus and reconstructing $x$ and $y$ locations along antigenic dimensions 1 and 2 backward through time.
We use MCMC to sample tip locations, but when outputting trees sample internal node locations using a peeling algorithm as described in [22].
Thus, after the MCMC is finished we have a posterior sample of 2000 trees each tagged with estimated tip locations and internal node locations.
We post-processed each posterior tree by conducting a linear interpolation between parent-child node locations to arrive at $x$ and $y$ values at intervals of 0.05 years for each virus.
Then, for each interval, $x$ and $y$ values are averaged across the sample of posterior trees.
We draw lines between these locations to approximate mean posterior diffusion paths.
As virus lineages coalesce backwards through time down the phylogeny these diffusion paths will also coalesce.''}

It would be helpful to have phylogenetic trees for the four clades as supporting figures, since as best I understand Figure 2 and 3 do not represent the phylogenetic relationships, but only show the influence of those relationships on the antigenic map.

\textbf{The virus phylogenies can be seen to some extent in Figures 2 and 3, i.e. phylogenetically distinct clades are apparent.  However, we agree that it would be helpful to include a cleaner picture of the virus phylogenies.  We've included a new figure (current figure 4) that shows more standard time-resolved phylogenies.  We've also included discussion of the negative correlation between antigenic drift and phylogenetic diversity.}

### 2013-07-24 flux.tex
Revise to note HA vs HA1 sequences.
Small reorganization.
Fix #17 - phylogenetic trees.

@@

\subsection*{Epidemiological consequences of antigenic drift}

@@

We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD. 
Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
However, the majority of isolates were present from 2002 to 2007. 
Because we are interested in longer-term antigenic evolution, we subsampled the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
This censoring left 402 virus isolates, 519 serum isolates and 10,059 HI measurements. 
Each virus isolate was compared to an average of 21.9 serum isolates, and each serum isolate was compared to an average of 18.0 virus isolates.

>>>>

These patterns of antigenic drift are reflected in the virus phylogenies (Figure~\ref{trees}).
Influenza A/H3N2 has characteristically spindly tree showing rapid turnover of the virus population, while A/H1N1 and B have trees that show greater degrees of viral coexistance (Figure~\ref{trees}).
The scaled effective population size $N_e \tau$ measures the timescale of coalescence for a phylogeny and quantifies the visual distinction between a `spindly' tree and a `bushy' tree \cite{Bedford11}.
Here, we see that $N_e \tau$ broadly correlates with the rate of antigenic drift with A/H3N2 showing fast drift and reduced effective population size (Figure~\ref{trees}), as expected from basic epidemiological models \cite{Bedford12}.
Antigenic drift results in the replacement of antigenically primitive lineages by antigenically advanced lineages, thereby reducing genealogical diversity.

%%% trees %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=0.85\textwidth]{figures/trees}
	\caption{\textbf{Time-resolved phylogenetic trees of A/H3N2, A/H1N1, B/Vic and B/Yam viruses.} 
	The maximium-clade credibility (MCC) tree is shown for each virus.
	These trees show genealogical relationships, so that branches are measured in terms of years rather than substitutions.
	The mean posterior scaled effective population size $N_e \tau$, measured in years, and its 95\% credible interval is shown for each virus.} 
	\label{trees} 
\end{figure}

\subsection*{Punctuated evolution and its epidemiological consequences}

@@

@@

We compiled an antigenic dataset of hemagglutination inhibition (HI) measurements of virus isolates against post-infection ferret sera for influenza A/H3N2 by collecting data from previous publications \cite{Hay01,Smith04,Russell08,Barr10}, NIMR vaccine strain selection reports for 2002 and 2008--2012 \cite{NIMR02,NIMRMarch08,NIMRFeb09,NIMRFeb10,NIMRSep10,NIMRSep11,NIMRFeb12} and the Feb 2011 VRBPAC report \cite{Cox11FDA}.
We queried the Influenza Research Database \cite{IRD} and the EpiFlu Database \cite{GISAID} for HA nucleotide sequences by matching strain names, e.g.\ A/HongKong/1/1968, and only strains for which sequence was present was retained.
If a strain had multiple sequences in the databases we preferentially kept the IRD sequence and preferentially kept the longest sequence in IRD.
Many strains had full length HA sequences, while other strains only possessed HA1 sequences.
Sequences were aligned using MUSCLE v3.7 under default parameters \cite{MUSCLE}.
This dataset had 2051 influenza isolates (present as either virus or serum in HI comparisons) dating from 1968 to 2011. 
However, the majority of isolates were present from 2002 to 2007. 
Because we are interested in longer-term antigenic evolution, we subsampled the data to have at most 20 virus isolates per year, preferentially keeping those isolates with more antigenic comparisons. 
We then kept only those serum isolates that are relatively informative to the antigenic placement of viruses, dropping serum isolates that are compared to 4 or fewer different virus isolates.
This censoring left 402 virus isolates, 519 serum isolates and 10,059 HI measurements. 
Each virus isolate was compared to an average of 21.9 serum isolates, and each serum isolate was compared to an average of 18.0 virus isolates.
